---
title: "Crim Patterns in Chicago"
author: "Paul Olulana"
date: "2024-12-12"
output:
  html_document:
    toc: true
    theme: cosmo
    self_contained: true
    code_folding: hide
    css: styles.css
   
---


#  {.tabset}

## <span style="color:blue; font-size:20px;">Introduction</span>



This report explores crime patterns in Chicago using a comprehensive analysis of 2015 data from the City of Chicago Open Data Portal. By leveraging R’s powerful data manipulation and visualization tools, the study examines temporal trends, such as variations in crime by month and hour, as well as geospatial patterns, with a focus on identifying violent crime hotspots across neighborhoods. The objective is to uncover actionable insights that can inform data-driven strategies for crime prevention and resource allocation, ultimately contributing to improved public safety and urban management.

<img src="https://encrypted-tbn0.gstatic.com/images?q=tbn:ANd9GcQX2uQm5FPUSKOIwuSgbyd4FsVFE9T6WHm20g&s" alt="Introduction Image" style="display:block; margin:auto; width:100%;">

---

## <span style="color:blue; font-size:20px;"> Data Preparation</span>

### Loading the Dataset
```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(sf)       
library(knitr)    
library(ggmap)    

# Set working directory 
setwd("C:/Users/paulo/OneDrive/Documents/BSB_Docs/R_Project/Final Individual Project/Paul_Olulana Final Project")

# Load dataset
crimes <- read.csv("Crimes_-_2015_20241212.csv")

# Explore the dataset structure
str(crimes)
head(crimes)

# Summarize missing data
missing_summary <- colSums(is.na(crimes))
kable(as.data.frame(missing_summary), col.names = c("Column", "Missing Values"))
```


### Data Cleaning 
```{r data-cleaning, echo=FALSE}
# Remove rows with missing values in critical columns
crimes <- crimes %>%
  filter(!is.na(Latitude) & !is.na(Longitude))

# Convert date to proper datetime format and extract useful components
crimes <- crimes %>%
  mutate(Date = mdy_hms(Date),
         Year = year(Date),
         Month = month(Date, label = TRUE, abbr = TRUE),
         Hour = hour(Date))

# Categorize crimes into broader categories
crimes <- crimes %>%
  mutate(Crime_Category = case_when(
    grepl("ASSAULT|BATTERY|HOMICIDE", Primary.Type, ignore.case = TRUE) ~ "Violent",
    grepl("BURGLARY|THEFT|ROBBERY", Primary.Type, ignore.case = TRUE) ~ "Property",
    TRUE ~ "Other"
  ))


```

## <span style="color:blue; font-size:20px;">Analysis and Visualization</span>

### Summarize Crime Categories
```{r crime-summary, echo=FALSE}
#Summarize total crimes by category
crime_summary <- crimes %>%
  group_by(Crime_Category) %>%
  summarise(Total = n()) %>%
  arrange(desc(Total))

kable(crime_summary, col.names = c("Crime Category", "Total Crimes"))
```

### Monthly Crime Trends
```{r monthly-trends, echo=FALSE}
crimes %>%
  group_by(Month) %>%
  summarise(Total = n()) %>%
  ggplot(aes(x = Month, y = Total, group = 1)) +
  geom_line(color = "blue", linewidth = 1) + 
  geom_point(color = "red", size = 2) +
  labs(title = "Monthly Crime Trends", x = "Month", y = "Total Crimes") +
  theme_minimal()

```

### Hourly Crime Distribution
```{r}
crimes %>%
  group_by(Hour) %>%
  summarise(Total = n()) %>%
  ggplot(aes(x = Hour, y = Total)) +
  geom_line(color = "green", linewidth = 1) +
  geom_point(color = "purple", size = 2) +
  labs(title = "Crimes by Hour", x = "Hour of Day", y = "Total Crimes") +
  theme_minimal()
```


### Mapping Crime Hotspots 
```{r crime-hotspots-heatmap, echo=FALSE}
# Filter for violent crimes
violent_crimes <- crimes %>% filter(Crime_Category == "Violent")

# Create a heatmap 
ggplot(violent_crimes, aes(x = Longitude, y = Latitude)) +
  stat_density2d(aes(fill = after_stat(level)), geom = "polygon", alpha = 0.7) + 
  scale_fill_viridis_c(option = "plasma") + 
  geom_point(alpha = 0.05, color = "black") + 
  labs(
    title = "Violent Crime Hotspots in Chicago (Heatmap)",
    x = "Longitude",
    y = "Latitude",
    fill = "Density Level"
  ) +
  theme_minimal()
```


### Analyzing Violent Crimes by Community Area
```{r neighborhood-analysis, echo=FALSE}
violent_by_area <- crimes %>%
  filter(Crime_Category == "Violent") %>%
  group_by(Community.Area) %>%
  summarise(Total = n()) %>%
  arrange(desc(Total))

#Display the table
kable(violent_by_area, col.names = c("Community Area", "Total Violent Crimes"))
```

 
### Top Crime Types
```{r top-crime-types, echo=FALSE}
crimes %>%
  count(Primary.Type, sort = TRUE) %>%
  top_n(10) %>%
  ggplot(aes(x = reorder(Primary.Type, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Crime Types", x = "Crime Type", y = "Count") +
  theme_minimal()
```

## <span style="color:blue; font-size:20px;">Conclusion</span>

### Conclussion 

This analysis of Chicago’s 2015 crime data revealed significant temporal and geospatial trends, with crime rates peaking during summer months and evening hours, and violent crimes concentrated in specific neighborhoods. Theft, battery, and narcotics-related offenses emerged as the most prevalent crime types. These insights highlight the need for targeted interventions, including increased resource allocation in high-crime areas, community engagement programs, and seasonal strategies to address peak crime periods. By demonstrating the value of data-driven approaches, this study underscores the potential of analytics in guiding effective public safety and urban policy decisions.
